Unsupervised Characterization of Water Composition with UAV-based Hyperspectral Imaging and Generative Topographic Mapping

### 2024-05-20 paper/paper.tex
updated w/ Adam's suggestions

Inland water bodies present a unique challenge to characterization by remote sensing imagery due to their complex spectral characteristics and small scale spatial variability. The broad bands of multi-spectral imagers coupled with the irregular shape of lakes and rivers results in pixels with highly mixed signals that are easily dominated by reflectance from shore and near-shore vegetation sources \cite{koponen2002lake, ritchie2003remote}. Recently, the combination of hyperspectral imaging with unmanned aerial vehicles (UAV) such as drones has emerged as a powerful approach to simultaneously address the spectral, spatial, and temporal limitations of traditional high-altitude and satellite-based collection \cite{adao2017hyperspectral,arroyo2019implementation}. UAV are significantly less expensive to deploy than their remote sensing counterparts, and low altitude flights enable centimeter-scale sampling while limiting the need for complicated atmospheric corrections \cite{banerjee2020uav}. However, the significant increase in data volume generated by these systems presents a new challenge, namely, how to efficiently extract water quality parameters of interest from intricate pixel spectra.

@@

In this paper, we explore the application of the GTM to UAV-acquired HSI for the characterization water quality. Using data collected at pond in Montague, North Texas, we first train a GTM using water-only pixels to produce a low-dimensional representation of the collected HSI. We use this mapping to explore the highly detailed small-scale variability within the pond and discuss how this can be used to guide reference data collection. Next, we demonstrate how the GTM can be utilized to identify relevant spectral endmembers from a combined dataset including land pixels, algae, water, and a simulated contaminant plume using a rhodamine tracer dye. Once identified, these endmembers can be used to rapidly map the abundance of spectral features within the pond. We demonstrate this capability by using GTM to map the abundance of algae near the shore as well as the dispersion of a rhodamine dye plume.

@@

The function $\psi$ is typically chosen to be given by the sum of $M$-many radial basis functions (RBF) evenly distributed in the latent space such that $\psi(\xi) = W\phi(\xi)$ with centers $\mu_m$ and width $\sigma$ so that
\begin{equation}
    \phi_m(\xi) = \exp\left(-\dfrac{\lVert \xi - \mu_m \rVert^2}{2\sigma^2}\right).
\end{equation}
The width $\sigma$ is taken to be the distance between neighboring RBF centers multiplied by a scale factor, $s$. to Together the number of RBFs, $M$, and $s$ are hyperparameters for the model which govern the smoothness of the resulting manifold in the data space. Additionally, sparsity of $W$ can be enforced by introducing an additional hyperparameter $\alpha$ corresponding to a prior distribution over the weights given by
\begin{equation}\label{eq:weight-prior}
    p(W \mid \alpha) =  \left( \frac{\alpha}{2\pi} \right)^{MD/2}\exp\left(-\frac{\alpha}{2}\left\lVert W \right\rVert_{F}^2\right).
\end{equation}

@@

Next, we consider a dataset of combined HSI pixels including both water and land. To simulate the dispersion of a potential contaminant source, a Rhodamine tracer dye was released into the western portion of the pond and two additional UAV flights were used to collect HSI capturing the evolution of the resulting plume. From these HSI a collection of $>145,000$ were sampled for model training. Exemplar spectra for water, grass, algae, and rhodamine dye were identified from a sample HSI as shown in panels (b) and (c) of Fig~\ref{fig:sample-spectra}. Using these spectra, we explore the trained GTM and use it to extract spectral signatures corresponding to these endmember categories. Values for the hyperparameters $m$, $s$, and $\alpha$, are determined by training multiple GTM models and selecting values which minimize the Bayesian Information Criterion (BIC) given by 
\begin{equation}
    \text{BIC} = 2P\ln(N) - 2\mathcal{L}
\end{equation}
where $P$ is the total number of model parameters, $N$ is the number of records in the dataset, and $\mathcal{L}$ is the log-likelihood defined in Eq~\ref{eq:llh}.

@@

A GTM with $K=32\times 32$ nodes was trained on the dataset of water-only pixels described in Sec~\ref{sec:case-studies} in order to explore the distribution of reflectance spectra captured by the HSI. The resulting GTM is visualized in Fig~\ref{fig:gtm-water}. First, we note that the GTM has utilized most of the embedding space as illustrated panel (b) where the position corresponding to the mean responsibility of each data point $\hat{\xi}_n$ has been plotted in the latent space. For this water-only gtm, the spectra appear to be largely clustered toward the left edge, bottom, and right edge of the embedding space. Spectral signatures corresponding to GTM nodes from the four corners and center of the latent space are shown in panel (c) illustrating the spectral variability represented across the latent space.

@@

A second GTM was trained on the combined dataset of HSI covering shore, water, and the rhodamine tracer dye release. The resulting class map is shown in Fig~\ref{fig:ndvi} where each sample has been plotted at the position of the mean node responsibility and colored by the normalized difference vegetation index (NDVI) computed from the original reflectance spectrum. The NDVI is a spectral index sensitive to variations in vegetation health with negative values corresponding to water, small positive values corresponding to sparse vegetation, and large values near $1$ corresponding to dense vegetation \cite{thenkabail2018hyperspectral}. From this, we see that the GTM can clearly separate spectra by vegetation content with high values concentrated to the left of the embedding space and negative, water-based pixels concentrated to the right. Additionally, the position of exemplar spectra for algae, rhodamine, water, and grass are included as color-filled circles further illustrating the classification obtained by the GTM.

@@

\begin{figure}[t]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=17.0cm]{paper/figures/results/algae.png}
\end{adjustwidth}
\caption{Using the spectral endmember assigned to algae from the trained GTM together with the NS3, we are able to estimate the algal abundance near the shore. (\textbf{left}) NS3 values of pixels below a threshold of 0.4275 corresponding to a total area of $587.3$ $m^2$. (\textbf{right}) A picture of the pond near the shore showing the algae.\label{fig:algae-map}}
\end{figure}

@@

The second application of the GTM presented in this study is for the extraction of endmembers corresponding to unique sources observed in the HSI. Here the GTM is an attractive choice as it does not rely on the assumption of linear mixing and the presence of pure pixels which are easily broken by the effects of multiple scattering in realistic scenes \cite{nonlinearity-in-hsi}. Additionally, because the GTM is a probabilistic model, the values of model hyperparameters can be selected objectively using information criteria like the BIC. This is a clear advantage over similar methods like the SOM on which the GTM is based. Endmembers identified for exemplar spectra corresponding to rhodamine, algae, grass, and water demonstrate that the method can successfully identify spectral signatures corresponding to diverse sources. If a set of labeled spectra for known sources are available, their representation in the GTM latent space could be used to perform as semi-supervised classification similar to the SOM approach developed by Riese et al. \cite{riese2019supervised}.

>>>>

Inland water bodies present a unique challenge to characterization by remote sensing imagery due to their complex spectral characteristics and small scale spatial variability. The broad bands of multi-spectral imagers coupled with the irregular shape of lakes and rivers results in pixels with highly mixed signals that are easily dominated by reflectance from shore and near-shore vegetation sources \cite{koponen2002lake, ritchie2003remote}. Recently, the combination of hyperspectral imaging with unmanned aerial vehicles (UAV) such as drones has emerged as a powerful approach to simultaneously address the spectral, spatial, and temporal limitations of traditional high-altitude and satellite-based collection \cite{adao2017hyperspectral,arroyo2019implementation}. UAVs are significantly less expensive to deploy than their remote sensing counterparts, and low altitude flights enable centimeter-scale sampling while limiting the need for complicated atmospheric corrections \cite{banerjee2020uav}. However, the significant increase in data volume generated by these systems presents a new challenge, namely, how to efficiently extract water quality parameters of interest from intricate pixel spectra.

@@

In this paper, we explore the application of the GTM to UAV-acquired HSI for the characterization water quality. Using data collected at a pond in Montague, North Texas, we first train a GTM using water-only pixels to produce a low-dimensional representation of the collected HSI. We use this mapping to explore the highly detailed small-scale variability within the pond and discuss how this can be used to guide reference data collection. Next, we demonstrate how the GTM can be utilized to identify relevant spectral endmembers from a combined dataset including land pixels, algae, water, and a simulated contaminant plume using a rhodamine tracer dye. Once identified, these endmembers can be used to rapidly map the abundance of spectral features within the pond. We demonstrate this capability by using a GTM to map the abundance of algae near the shore as well as the dispersion of a rhodamine dye plume.

@@

The function $\psi$ is typically chosen to be given by the sum of $M$-many radial basis functions (RBF) evenly distributed in the latent space such that $\psi(\xi) = W\phi(\xi)$ with centers $\mu_m$ and width $\sigma$ so that
\begin{equation}
    \phi_m(\xi) = \exp\left(-\dfrac{\lVert \xi - \mu_m \rVert^2}{2\sigma^2}\right).
\end{equation}
The width $\sigma$ is taken to be the distance between neighboring RBF centers multiplied by a scale factor, $s$. Together the number of RBFs, $M$, and $s$ are hyperparameters for the model which govern the smoothness of the resulting manifold in the data space. Additionally, sparsity of $W$ can be enforced by introducing an additional hyperparameter $\alpha$ corresponding to a prior distribution over the weights given by
\begin{equation}\label{eq:weight-prior}
    p(W \mid \alpha) =  \left( \frac{\alpha}{2\pi} \right)^{MD/2}\exp\left(-\frac{\alpha}{2}\left\lVert W \right\rVert_{F}^2\right).
\end{equation}

@@

Next, we consider a dataset of combined HSI pixels including both water and land. To simulate the dispersion of a potential contaminant source, a Rhodamine tracer dye was released into the western portion of the pond and two additional UAV flights were used to collect HSI capturing the evolution of the resulting plume. From these HSI a collection of $>145,000$ pixels were sampled for model training. Exemplar spectra for water, grass, algae, and rhodamine dye were identified from a sample HSI as shown in panels (b) and (c) of Fig~\ref{fig:sample-spectra}. Using these spectra, we explore the trained GTM and use it to extract spectral signatures corresponding to these endmember categories. Values for the hyperparameters $m$, $s$, and $\alpha$, are determined by training multiple GTM models and selecting values which minimize the Bayesian Information Criterion (BIC) given by 
\begin{equation}
    \text{BIC} = 2P\ln(N) - 2\mathcal{L}
\end{equation}
where $P$ is the total number of model parameters, $N$ is the number of records in the dataset, and $\mathcal{L}$ is the log-likelihood defined in Eq~\ref{eq:llh}.

@@

A GTM with $K=32\times 32$ nodes was trained on the dataset of water-only pixels described in Sec~\ref{sec:case-studies} in order to explore the distribution of reflectance spectra captured by the HSI. The resulting GTM is visualized in Fig~\ref{fig:gtm-water}. First, we note that the GTM has utilized most of the embedding space as illustrated panel (b) where the position corresponding to the mean responsibility of each data point $\hat{\xi}_n$ has been plotted in the latent space. For this water-only GTM, the spectra appear to be largely clustered toward the left edge, bottom, and right edge of the embedding space. Spectral signatures corresponding to GTM nodes from the four corners and center of the latent space are shown in panel (c) illustrating the spectral variability represented across the latent space.

@@

A second GTM was trained on the combined dataset of HSI covering shore, water, and the rhodamine tracer dye release. The resulting class map is shown in Fig~\ref{fig:gtm-ndvi} where each sample has been plotted at the position of the mean node responsibility and colored by the normalized difference vegetation index (NDVI) computed from the original reflectance spectrum. The NDVI is a spectral index sensitive to variations in vegetation health with negative values corresponding to water, small positive values corresponding to sparse vegetation, and large values near $1$ corresponding to dense vegetation \cite{thenkabail2018hyperspectral}. From this, we see that the GTM can clearly separate spectra by vegetation content with high values concentrated to the left of the embedding space and negative, water-based pixels concentrated to the right. Additionally, the position of exemplar spectra for algae, rhodamine, water, and grass are included as color-filled circles further illustrating the classification obtained by the GTM.

@@

\begin{figure}[t]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=17.0cm]{paper/figures/results/algae.png}
\end{adjustwidth}
\caption{Using the spectral endmember assigned to algae from the trained GTM together with the NS3, we are able to estimate the algal abundance near the shore. (\textbf{Top}) NS3 values of pixels below a threshold of 0.4275 corresponding to a total area of $587.3$ $m^2$. (\textbf{Bottom}) A picture of the pond near the shore showing the algae.\label{fig:algae-map}}
\end{figure}

@@

The second application of the GTM presented in this study is for the extraction of endmembers corresponding to unique sources observed in the HSI. Here the GTM is an attractive choice as it does not rely on the assumption of linear mixing and the presence of pure pixels which are easily broken by the effects of multiple scattering in realistic scenarios \cite{nonlinearity-in-hsi}. Additionally, because the GTM is a probabilistic model, the values of model hyperparameters can be selected objectively using information criteria like the BIC. This is a clear advantage over similar methods like the SOM on which the GTM is based. Endmembers identified for exemplar spectra corresponding to rhodamine, algae, grass, and water demonstrate that the method can successfully identify spectral signatures corresponding to diverse sources. If a set of labeled spectra for known sources are available, their representation in the GTM latent space could be used to perform as semi-supervised classification similar to the SOM approach developed by Riese et al. \cite{riese2019supervised}.
